Retrofitting Effect Handlers onto OCaml

### 2020-10-28 retro-concurrency.tex
minor typo readability edits

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis such as debuggers and profilers that utilise DWARF stack unwind tables
to obtain a backtrace. Our primary motivation for adding effect handlers to
OCaml is to support (1) highly concurrent applications such as web servers to
be expressed naturally in \emph{direct-style} as opposed to using callbacks and
(2) efficient data and task parallel programs in the vein of OpenMP~\cite{} and
Intel TBB~\cite{}.

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging for implement delimited
continuations necessary for effect handlers without sacrificing the desirable
properties of OCaml. One standard way of implementing continuations is to use
continuation-passing style (CPS) in the compiler's intermediate representation
(IR)~\cite{Koka}. If the compiler were to use CPS, then reifying continuations
in the source language is relatively straight-forward. But OCaml does not use
CPS IR, and changing the OCaml compiler to utilise CPS IR would be an enormous
undertaking. CPS IR would also affect the performance profile of OCaml
applications due to increased memory allocations as continuation closures get
allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding in
meaningless and hence, we would lose compatibility with tools that inspect the
program stack. Hence, we choose not to CPS translation.

@@

We have implemented effect handlers in a multicore extension of the OCaml
programming language. We call our implementation ``Multicoe OCaml'' to
distinguish it from stock OCaml. Multicore OCaml uses effect handlers as the
mechanism to express user-level concurrency.

>>>>

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with fast
foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables to obtain a backtrace. Our primary motivation for adding effect handlers
to OCaml is to support (1) highly concurrent applications such as web servers to
be expressed naturally in \emph{direct-style} as opposed to using callbacks and
(2) efficient data and task parallel programs in the vein of OpenMP~\cite{} and
Intel TBB~\cite{}.

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
continuations in the source language is relatively straight-forward. But OCaml
does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
an enormous undertaking. CPS IR would also affect the performance profile of
OCaml applications due to increased memory allocations as continuation closures
get allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding
is meaningless and hence, we would lose compatibility with tools that inspect
the program stack. Hence, we choose not to CPS translation.

@@

We have implemented effect handlers in a multicore extension of the OCaml
programming language. We call our implementation ``Multicore OCaml'' to
distinguish it from stock OCaml. Multicore OCaml uses effect handlers as the
mechanism to express user-level concurrency.
